---
title: "Eval expressions"
output:
    html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/package/
  calls_path: summarized.fst
  corpus_path: ../corpus.fst
  static_path: evals-static.csv
  force_rebuild: TRUE
---

```{r setup, include=FALSE}
library(tidyverse)
library(fst)
library(fs)
library(DT)
library(scales)

knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) { # record the current time before each chunk
      now <<- Sys.time()
    } else { # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("Time for this code chunk to run:", round(res, 2), units(res))
    }
  }
}))
now <- Sys.time()
knitr::opts_chunk$set(
  echo = TRUE,
  fig.retina = 2,
  fig.width = 10,
  cache.lazy = FALSE,
  time_it = TRUE
)

source("inc/paths.R")
source("inc/latextags.R")
source("inc/functions.R")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)

calls_path <- paste0(params$base_dir, params$calls_path)
corpus_path <- paste0(params$base_dir, params$corpus_path)
static_path <- paste0(params$base_dir, params$static_path)
```



# Read

Read in the summarized calls data (`E_raw`) and the file that describes
which package belong to our corpus (`C_raw`). For conciseness, `E` is
the input data set with only the columns that we use.

```{r read, message=TRUE, include=F}
read_fst(corpus_path) %>% as_tibble() -> C_raw
read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P

read_fst(calls_path) %>% as_tibble() -> E_raw

E_raw %>% select(
  run_package = package, #      The package which was run to trigger the eval
  ev_variant = eval_function, # Which eval variant was called
  duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
  src_ref = eval_call_srcref, # Source ref for the eval call site
  file, #                       Script file name (name of this run)
  ev_package = eval_source, #   Package of the call site
  type = expr_resolved_type_tag, #   Argument type
  ev_expr = expr_expression,
  expr_resolved,
  func = expr_resolved_function, # the function (if any) at the root of the AST of expr_resolved
  parsed = expr_parsed_expression
) -> E

C_raw -> C
corpus_size <- nrow(C)
```


Some useful numbers:
```{r}
nb_call_sites <- E %>%
  pull(src_ref) %>%
  n_distinct()
```


# Most frequent expressions


Before resolving, so as the programmer wrote them:

```{r}
E %>% count(ev_expr, wt = duplicates, sort = TRUE)
```


After resolving:

```{r}
E %>% count(expr_resolved, wt = duplicates, sort = TRUE)
```


# Resolved expressions


Are expressions resolved to very diverse values or expressions?
If they are uniquely resolved and it is a value, then we can get rid of the `eval`.

```{r}

resolved_variety <- E %>%
  group_by(ev_expr, src_ref) %>%
  summarise(n = n_distinct(expr_resolved)) %>%
  arrange(desc(n))
resolved_variety %>% datatable()

nb_one_resolved <- resolved_variety %>%
  filter(n == 1) %>%
  nrow()
```

There are `r nb_one_resolved` call sites that are resolved to only one expression, out of `r nrow(resolved_variety)`.

We can now count how many of these call sites with only one resolved expression resolve to a value type.

```{r}
one_resolved_type <- E %>%
  group_by(src_ref) %>%
  mutate(n_resolved = n_distinct(expr_resolved)) %>%
  filter(n_resolved == 1) %>%
  ungroup() %>%
  select(src_ref, n_resolved, type) %>%
  distinct()

nb_one_resolved_type <- one_resolved_type %>% nrow()
```


`r nb_one_resolved_type` call sites resolved to one expression, of only one type.

```{r}
nb_value_type <- one_resolved_type %>% filter(type %in% c("character", "double", "integer", "logical", "NULL"))
```


There are `r nb_value_type %>% nrow()` such call sites.


# Expressions common to several call sites and packages

It emphasizes which expressions are a common pattern across various code bases.

```{r}
widespread_expr <- E %>%
  group_by(expr_resolved) %>%
  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
```


```{r}
widespread_expr %>%
  arrange(desc(n_call_sites)) %>%
  datatable()
```

```{r}
widespread_expr %>%
  arrange(desc(n_packages)) %>%
  datatable()
```



For instance, for the most widespread expression, both in terms of call sites and packages, `mf`, we show here the packages sorted by decreasing number of call sites with this expression name.

```{r}
E %>%
  filter(ev_expr == "mf") %>%
  group_by(ev_package) %>%
  summarise(n = n_distinct(src_ref)) %>%
  arrange(desc(n))
```


Manual inspection seems to show that `mf` is often used as a result of a `match.call`:

- https://github.com/cran/PMCMRplus/blob/4f5234542aa0a60642b230bea7efb668fb16a140/R/mandelkTest.R#L121 
- https://github.com/fbertran/plsRglm/blob/dfa4e54ea02bca8bf04d29bb65dc7dba611927c9/R/plsRglmmodel.default.R#L4
- https://github.com/cran/RWeka/blob/55030d5513bf37760d1f6e2f1a49495977113276/R/classifiers.R#L42 with a comment _## The "usual" way of creating a model frame from the call._ (sic!)

This is the common pattern when a call is retrieved and then modified, then evaluated.

We can also look at the functions called in the expression:

```{r}
widespread_function <- E %>%
  filter(!is.na(func)) %>%
  group_by(func) %>%
  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
```


# Expressions using `parse`, `str2lang`, and `str2expression`

```{r}
parsed_expr <- E %>% filter(!is.na(parsed))

nb_parsed_call_sites <- parsed_expr %>%
  pull(src_ref) %>%
  n_distinct()
```


There are `r nb_parsed_call_sites` call sites using those functions in their argument, out of the `r nb_call_sites` in total, i.e. `r ratio(nb_parsed_call_sites, nb_call_sites)`%.

# Arithmetic expressions

First, we only keep rows where the root is an arithmetic function, and the rows with a constant number

```{r}
arith_func <- c("/", "-", "*", "+", "^", "log", "sqrt", "exp", "max", "min", "cos", "sin", "abs", "atan", ":")

arithm <- E %>% filter(func %in% arith_func)
```

We want to keep only the ones that only use these functions and constants.

```{r}
is_arithm_expr <- function(exp) {
  if (is.call(exp)) {
    function_name <- exp[[1]]
    function_args <- exp[-1]
    return(as.character(function_name) %in% arith_func && every(function_args, is_arithm_expr))
  }
  else if (is.expression(exp)) {
    return(every(exp, is_arithm_expr))
  }
  else if (typeof(exp) %in% c("integer", "double", "complex")) {
    return(TRUE)
  }
  else {
    return(FALSE)
  }
}

arithm_folded <- arithm %>% filter(map_lgl(expr_resolved, function(exp) {
  ast <- NA
  try(ast <- parse(text = exp)[[1]], silent = TRUE)
  # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
  if (length(ast) > 1 || !is.na(ast)) {
    return(is_arithm_expr(ast))
  }
  else {
    return(FALSE)
  }
}))
```

It corresponds to these call sites and packages:

```{r}
n_arithm_call_sites <- arithm_folded %>%
  pull(src_ref) %>%
  n_distinct()

arithm_folded %>%
  group_by(expr_resolved) %>%
  summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>%
  datatable()
```


The total number of call sites with those arithmetic on constants is `r n_arithm_call_sites`.

# Formulas

A formula is defined with the tilde `~` operator. 

```{r}
with_formula <- E %>%
  filter(str_detect(expr_resolved, fixed("~")))
```

```{r}
n_with_formula <- with_formula %>%
  pull(src_ref) %>%
  n_distinct()
```


The total number of call sites with formulas is `r n_with_formula`.

```{r}
with_formula %>%
  group_by(expr_resolved) %>%
  summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>%
  datatable()
```

`~` is mostly used within a call to `model.frame`.


# Variables

Here, we analyze expressions with only a variable name.

```{r}
with_var <- E %>%
  filter(str_detect(expr_resolved, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$")) %>%
  filter(expr_resolved != "NA") # "NA" matches the regex of a variable name but is a value so we get rid of it
```


We can categorize them by number of call sites and number of packages:
```{r}
n_with_var <- with_var %>%
  pull(src_ref) %>%
  n_distinct()

with_var %>%
  group_by(expr_resolved) %>%
  summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>%
  datatable()
```

There are `r n_with_var` such call sites.


# Better normalization of resolved expressions

We fold all constant arithmetic expressions and replace all variable names by a placeholder.

```{r}
str_op <- c("paste", "paste0", "str_c")
comp_op <- c("<", ">", "<=", ">=", "==", "!=")
bool_op <- c("&", "&&", "|", "||", "!")

canonic_expr <- function(exp, with.names = FALSE) {
    if (is.call(exp)) {
        function_name <- exp[[1]]
        function_args <- exp[-1]
        res <- map(function_args, canonic_expr, with.names) # TODO: rather directly compute canonic_expr on function_args?
        # Function_name can be larger than one is a function with a namespace stats::model or another higher-order function
        # For instance, (function(x, y) x + y)(3, 4)
        if (length(function_name) == 1 && as.character(function_name) == "(") {
            return(res[[1]]) # (a) => (a)()
        }
        else if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
            v == "NUM"
        })) {
            return("NUM")
        }
        else if (length(function_name) == 1 && as.character(function_name) %in% str_op && every(res, function(v) {
            v == "STR"
        })) {
            return("STR")
        }
        else if (length(function_name) == 1 && as.character(function_name) %in% bool_op && every(res, function(v) {
            v == "BOOL"
        })) {
            return("BOOL")
        }
        else if (length(function_name) == 1 && as.character(function_name) %in% comp_op && every(res, function(v) {
            v == "NUM"
        })) {
            return("BOOL")
        }
        else if (length(function_name) == 1 && as.character(function_name) %in% c("c", "list") && length(res) > 1 && n_distinct(res) == 1 && res[[1]] %in% c("BOOL", "NUM", "STR", "VAR")) {
            str_c(as.character(function_name), "(", res[[1]], ")")
        }
        else if (length(function_name) == 1 && function_name == "::") { # we drop the namespace
            as.character(function_args[[2]])
        }
        else {
            function_name <- if (length(function_name) == 1) {
                func_name <- as.character(function_name)
                function_name <- if (func_name %in% arith_func) {
                    "OP"
                }
                else if (func_name %in% comp_op) {
                    "COMP"
                }
                else if (func_name %in% bool_op) {
                    "LOGI"
                }
                else {
                    func_name
                }
            }
            else {
                canonic_expr(function_name, with.names)
            }

            # It might have been transformed by canonic_expr in the recursive call above
            # it should always have length one
            res <- if (length(function_name) == 1) {
                func_name <- as.character(function_name)
                # VAR is absorbing: if there is at least one, we propagate to all of them
                res <- if (func_name %in% c("COMP", "LOGI", "OP") && "VAR" %in% res) {
                    rep.int("VAR", length(res))
                }
                else if (func_name == "model.frame") { # special case
                    # We keep the 1st two arguments (usually formula and data)
                    # and subset if it is there
                    subset_arg <- if ("subset" %in% names(res)) {
                        subset_s <- as.character(res["subset"])
                        res["subset"] <- NULL
                        str_c("subset = ", subset_s)
                    }
                    else {
                        "subset = NULL"
                    }
                    first_args <- if(length(res) > 1) {
                          res[1:2]
                    }
                    else {
                      c(res[[1]], "NULL")
                    }
                  c(first_args, subset_arg)
                }
                else {
                    res
                }
            }
            
            fun_args <- if(with.names) {
              str_c(names(res), if_else(names(res) == "", "",  " = "), res, collapse = ", ")
            } 
            else
            {
              str_c(res, collapse = ", ")
            }

            return(str_c(str_c(function_name, collapse = ", "), "(", fun_args, ")"))
        }
    }
    else if (is.symbol(exp)) {
        return("VAR")
    }
    else if (is.expression(exp)) {
        return(str_c(map(exp, canonic_expr, with.names), collapse = ", "))
    }
    else if (typeof(exp) %in% c("integer", "double", "complex")) {
        return("NUM")
    }
    else if (typeof(exp) == "logical") {
        return("BOOL")
    }
    else if (typeof(exp) == "character") {
        return("STR")
    }
    else {
        return(deparse1(exp))
    }
}

canonic_expr_str <- function(exp, with.names = FALSE) {
    ast <- NA
    try(ast <- parse(text = exp)[[1]], silent = TRUE)
    # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
    if (is.symbol(ast) || is.language(ast) || length(ast) > 1 || !is.na(ast)) {
        return(canonic_expr(ast, with.names))
    }
    else {
        return("NORMALIZATION ERROR")
    }
}

canonic_expr_str2 <- function(exp, with.names = FALSE) {
    ast <- NA
    try(ast <- parse(text = exp)[[1]], silent = TRUE)
  return("")
}
```

For performance reason, we only apply this function to the unique resolved expressions.

```{r}
res_exprs_uniq <- E %>%
  select(expr_resolved) %>%
  unique()

res_exprs <- res_exprs_uniq %>% mutate(expr_canonic = map_chr(expr_resolved, canonic_expr_str))

# res_exprs <- res_exprs %>% mutate(expr_canonic_names = map_chr(expr_resolved, canonic_expr_str), with.names = TRUE)

E_cano <- E %>% left_join(res_exprs)
```


```{r}
widespread_canonic <- E_cano %>%
  group_by(expr_canonic) %>%
  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))

widespread_canonic %>%
  arrange(desc(n_call_sites)) %>%
  datatable()
```
